Wait-free Queues with Polygarithmic Step Complexity

### 2022-08-15 article/psuedocode.tex
review ordering,get

@@

\begin{itemize}
\item We call operations strictly before $op$ in the sequence of operations $S$, prefix of the $op$.
  \item $E(n,b)$ is the sequence of enqueue operations $\in$ \nf{ops(n.blocks[b])} ordered by their process id.
  \item $E_{n,b,i}$ is the $i$th enqueue in $E(n,b)$.
  \item $D(n,b)$ is the sequence of dequeue operations $\in$ \nf{ops(n.blocks[b])} ordered by their process id.
  \item $D_{n,b,i}$ is the $i$th enqueue in $D(n,b)$.
\item Order of the enqueue operations in $n$: $E(n)=E(n,1).E(n,2).E(n,3)...$
\item $E_{n,i}$ is the $i$th enqueue in $E(n)$.
\item Order of the dequeue operations in $n$: $D(n)=D(n,1).D(n,2).D(n,3)...$
\item $D_{n,i}$ is the $i$th dequeue in $D(n)$.
\item Linearization: $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
\end{itemize}
\end{definition}
\it{Note that in the non-root nodes we only order enqueues and dequeues among the operations of their own type. Since \nf{GetENQ()} only searches among enqueues and \nf{IndexDEQ()} works on dequeues.}

\begin{lemma}[Get correctness] \label{get}
If \nf{n.blocks[b].num\sub{enq}$\geq$i} then \nf{n.GetENQ(b,i)} returns $E_{n,b,i}$.
\end{lemma}
\begin{proof}
We are going to prove this lemma by induction on the height of node \nf{n}. The base case, where \nf{n} os a leaf, is straightforward (Line \ref{getBaseCase}). Leaf blocks each contain exactly one operation, so only \nf{n.GetENQ(b,1)} can be called where \nf{n} is a leaf. \nf{n.GetENQ(b,1)} returns the operation stored in the $b$th block of leaf \nf{n}.

For the induction step we prove \nf{n.GetENQ(b,i)} returns $E_{n,b,i}$, if \nf{n.child.GetENQ(b,i)} returns $E_{n.child,b,i}$. For non-leaf nodes it is decided that the \nf{i}th enqueue in block \nf{b} of internal node \nf{n} is in the \nf{n.blocks[b]}'s subblocks in the left child of \nf{n} or in the \nf{n.blocks[b]}'s subblocks in the right child of \nf{n} (line \ref{leftOrRight}). From Definition \ref{ordering} we know enqueue operations in a block are ordered by their process id and since in leaves of the tree are ordered by process id from left to right, thus operations from the left subblocks come before operations from the right subblocks in a block (See Figure \ref{figGet}). Furthermore \nf{b.num\sub{enq-left}} stores the number of \nf{enqueue()} operations from the \nf{n.blocks[b]}'s subblocks in the left child of \nf{n}. So if \nf{i} is greater than \nf{b.num\sub{enq-left}} it means $i$th operation is propagated from the right child, otherwise we should search for the $i$th enqueue in the left child. By definition \ref{def::ops} and \ref{def::subblock} we need to search in subblocks of \nf{n.blocks[b]} from the range \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.

If the $i$th \nf{enqueue} of \nf{n.blocks[b]} is in the left child it would be $i$th enqueue in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} by Definition \ref{def::subblock}. Also we know there are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before this range, so $E_{n,b,i}$ is $E_{n.left, i+eb}$, which is $E_{n.left,b^\prime,i^\prime}$ for some $b^\prime$ and $i^\prime$. We can compute $b^\prime$ search for $i+eb$th enqueue in \tt{n.left} and $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}. The parameters in \ref{leftChildGet} are for searching $E_{n.left, i+eb}$ in \nf{n.left.block} in the expected range of blocks, so this \nf{BSearch} returns the index of the subblock containing $E_{n,b,i}$.

@@

  \caption{The number and ordering of the enqueues propagated from the left and the right child to \nf{n.blocks[b]}. Enqueue operations from the left subblocks, colored red, are ordered before the enequeue operations from the right child, colored blue.
  }\label{figGet}
\end{figure}

@@

\begin{figure}[hbt]  
  \center\includegraphics[width=3in]{pics/end-b.png}

  \caption{The number written on top of the bars is the queue size. the first block is $b$ and the last block is $end$.}
  \label{end-b}
\end{figure}

>>>>

\begin{itemize}
\item Prefix of the $op$ in the sequence of operations $S$, is the operations strictly before $op$.
  \item $E(n,b)$ is the sequence of enqueue operations in \nf{ops(n.blocks[b])} ordered by their process id.
  \item $E_{n,b,i}$ is the $i$th enqueue in $E(n,b)$.
  \item $D(n,b)$ is the sequence of dequeue operations in \nf{ops(n.blocks[b])} ordered by their process id.
  \item $D_{n,b,i}$ is the $i$th enqueue in $D(n,b)$.
\item Order of the enqueue operations in the node $n$: $E(n)=E(n,1).E(n,2).E(n,3)...$
\item $E_{n,i}$ is the $i$th enqueue in $E(n)$.
\item Order of the dequeue operations in the node $n$: $D(n)=D(n,1).D(n,2).D(n,3)...$
\item $D_{n,i}$ is the $i$th dequeue in $D(n)$.
\item Linearization: $L=E(root,1).D(root,1).E(root,2).D(root,2).E(root,3).D(root,3)...$
\end{itemize}
\end{definition}
\it{Note that in the non-root nodes we only need ordering of enqueues and dequeues among the operations of their own type. Since \nf{GetENQ()} only searches among enqueues and \nf{IndexDEQ()} works on dequeues.}

@@

\begin{lemma}[Get correctness] \label{get}
If \nf{n.blocks[b].num\sub{enq}$\geq$i} then \nf{n.GetENQ(b,i)} returns $E_{n,b,i}$.
\end{lemma}
\begin{proof}
We are going to prove this lemma by induction on the height of node \nf{n}. The base case, where \nf{n} is a leaf, is straightforward (Line \ref{getBaseCase}). Leaf blocks each contain exactly one operation, so only \nf{n.GetENQ(b,1)} can be called where \nf{n} is a leaf. \nf{n.GetENQ(b,1)} returns the \nf{element} of the \nf{enqueue} operation stored in the $b$th block of leaf \nf{n}.

For the induction step we prove \nf{n.GetENQ(b,i)} returns $E_{n,b,i}$, if \nf{n.child.GetENQ(b,i)} returns $E_{n.child,b,i}$. In the Line\ref{leftOrRight} it is decided for the non-leaf nodes that the \nf{i}th enqueue in  \nf{b}th block  of internal node \nf{n} is in the \nf{n.blocks[b]}'s left child or right child subblocks. From Definition  \ref{ordering} of $E(n,b)$ we know enqueue operations in a block are ordered by their process id and since the leaves of the tree are ordered by process id from left to right, thus operations from the left subblocks come before operations from the right subblocks in a block (See Figure \ref{figGet}). Furthermore the \nf{num\sub{enq-left}} field  in a block stores the number of \nf{enqueue()} operations from the blocks's subblocks in the left child of \nf{n}. So $i$th enqueue operation is propagated from the right child if \nf{i} is greater than \nf{b.num\sub{enq-left}}. otherwise we should search for the $i$th enqueue in the left child. By definition \ref{def::ops} and \ref{def::subblock} we need to search in subblocks of \nf{n.blocks[b]} from the range \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} $\cup$ \texttt{n.right.blocks[n.blocks[i-1].end\textsubscript{right}+1..n.blocks[i].end\textsubscript{right}]}.

If the $i$th \nf{enqueue} of \nf{n.blocks[b]} is in the left child it would be $i$th enqueue in \texttt{n.left.blocks[n.blocks[i-1].end\textsubscript{left}+1..n.blocks[i].end\textsubscript{left}]} by Definition \ref{def::subblock}. Also we know there are $eb=n.blocks[b-1].sum_{enq-left}$ enqueues in the blocks before this range, so $E_{n,b,i}$ is $E_{n.left, i+eb}$ which is $E_{n.left,b^\prime,i^\prime}$ for some $b^\prime$ and $i^\prime$. We can compute $b^\prime$ search for $i+eb$th enqueue in \tt{n.left} and $i^\prime$ is \tt{i+eb-n.left.blocks[}$b^\prime-1$\tt{].sum\sub{enq}}. The parameters in \ref{leftChildGet} are for searching $E_{n.left, i+eb}$ in \nf{n.left.block} in the expected range of blocks, so this \nf{BSearch} returns the index of the subblock containing $E_{n,b,i}$.

@@

  \caption{The number and ordering of the enqueue operations propagated from the left and the right child to \nf{n.blocks[b]}. Enqueue operations from the left subblocks (colored red), are ordered before the enqueue operations from the right child (colored blue).
  }\label{figGet}
\end{figure}

@@

%\begin{figure}[hbt]  
%  \center\includegraphics[width=3in]{pics/end-b.png}
%
%  \caption{The number written on top of the bars is the queue size. the first block is $b$ and the last block is $end$.}
%  \label{end-b}
%\end{figure}
